Fire danger: the skill provided by the ECMWF Integrated Forecasting System

### 2019-12-23 article/article.Rmd
Article stage after calculating CRPS and extracting data for POD

# Load relevant packages
library("caliver")
library("raster")
library("ggplot2")
library("verification")
library("rowr")
library("dplyr")
library("rgeos")
library("maptools")
library("sp")
library("RColorBrewer")
library("stringr")
library("colorspace")
```

```{r areas, echo = FALSE, message = FALSE, results = 'hide', cache = TRUE}
## Get GFED4 regions and countries of interest
dfgfed4 <- data.frame(ID = 1:14,
                      Region = c("BONA", "TENA", "CEAM", "NHSA", "SHSA", "EURO",
                                 "MIDE", "NHAF", "SHAF", "BOAS", "CEAS", "SEAS",
                                 "EQAS", "AUST"),
                      Zone = c("north", "north", "tropics", "tropics", "south",
                               "north", "north", "tropics", "south", "north",
                               "north", "tropics", "tropics", "south"),
                      stringsAsFactors = FALSE)

BasisRegions <- readRDS("../data/BasisRegions_simplified.rds")

# GFED4 regions of interest
EURO <- BasisRegions[BasisRegions$Region == "EURO",]
TENA <- BasisRegions[BasisRegions$Region == "TENA",]
SHSA <- BasisRegions[BasisRegions$Region == "SHSA",]
# Countries of interest
pt <- raster::getData(name = "GADM", country = "Portugal", level = 1)[-c(2,13),]
chile <- raster::getData(name = "GADM", country = "Chile", level = 0)
cali <- raster::getData(name = "GADM", country = "USA", level = 1)[5, ]

# All the GFED4 regions
BasisRegions_join <- BasisRegions
# add id column for join
BasisRegions_join@data$id <- 1:nrow(BasisRegions_join@data)
regionXFort <- fortify(BasisRegions_join, data = BasisRegions_join@data)
regionXPoly <- merge(regionXFort, BasisRegions_join@data,
                     by.x = "id", by.y = "id")  # join data
# Re-order factors
regionXPoly$Region <- factor(as.character(regionXPoly$Region),
                             levels = dfgfed4$Region)
```

```{r Figure1, echo = FALSE, fig.width = 13, fig.height = 6, out.width = "100%", fig.cap = "\\label{fig:Figure1}GFED4 regional classification and the 3 countries selected to showcase the fire forecast performances (California, Chile, and Portugal)."}
ggplot(aes(x = long, y = lat), data = map_data("world")) +
  geom_polygon(aes(group = group), fill=NA, colour = "grey65") +
  coord_equal() +  theme_bw() + xlab("Longitude") + ylab("Latitude") +
  geom_polygon(data = regionXPoly,
               aes(x = long, y = lat, group = group, fill = factor(Region)),
               alpha = 0.5) +
  scale_fill_manual(name="GFED4 regions",
                    values = c("darkgrey",
                               RColorBrewer::brewer.pal(12,"Paired"),
                               "lightgrey"),
                    labels = dfgfed4$Region) +
  geom_polygon(data = map_data("world", region = "Portugal"),
               aes(x = long, y = lat, group = group, colour = "Portugal"),
               fill = NA) +
  geom_polygon(data = map_data("world", region = "Chile"),
               aes(x = long, y = lat, group = group, colour = "Chile"),
               fill = NA) +
  geom_polygon(data = map_data(map = "state", region = "California"),
               aes(x = long, y = lat, group = group, colour = "California"),
               fill = NA) +
  scale_colour_manual(name = "Study areas",
                      values = c("Portugal" = "brown",
                                 "California" = "darkblue",
                                 "Chile" = "darkgreen"))
```

@@

```{r Figure2, echo = FALSE, fig.width = 13, fig.height = 6, out.width = "100%", fig.cap = "\\label{fig:Figure2}Spatial distribution of weather stations from the synop network  which have at least 30 observations recorded at local noon in 2017."}
# Get data from previous publication
# https://github.com/cvitolo/GEFF-ERA5/blob/master/data/df_all.rds
# save them in the "../data" folder
# Use QGIS to beautify the map
knitr::include_graphics("../images/synops_col.png")
```

```
\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{station_error.png}
    \caption{Comparison between modelled FWI and observed FWI value. FWI are calculated from using ECMWF reanalysis (ERA5) and forecasts at different lead times. The box plots are used to describe the distribution of values accross the observation points for the 2017 Mean Bias  (panel a) and the 2017 mean absolute error (panel b)} \label{fig:synop}
\end{figure*}
```

Despite its importance  the analysis performed using the synop network it is still pointwise and does not cover all the regions where fires are relevant. Moreover mean biases and MAE are based on the high resolution forecast. These skill metrics do not provide information about the performance of the ensemble forecasting as a whole.

Figure \ref{fig:crp}

\textbf{ PUT here crps skills scores}

\subsection{ Skill in detecting fire events}

@@

>>>>

# Load relevant packages
library("caliver")
library("raster")
library("ggplot2")
library("verification")
library("rowr")
library("dplyr")
library("rgeos")
library("maptools")
library("sp")
library("RColorBrewer")
library("stringr")
library("colorspace")
library("easyVerification")
library("spatialEco")

forecasts_folder <- "/hugetmp/forecasts/fwi"

dates2017 <- seq.Date(from = as.Date("2017-01-01"),
                      to = as.Date("2017-12-31"),
                      by = "day")

datesERA5 <- seq.Date(from = as.Date("1980-01-01"),
                      to = as.Date("2019-06-30"),
                      by = "day")

# Get all GFED4 regions
BasisRegions <- readRDS("data/BasisRegions_simplified.rds")
```

@@

```{r prep, echo = FALSE, message = FALSE, results = 'hide', cache = TRUE}
# The following code uses forecast data which is restricted.
# A summary table is computed and saved for users at the end of this code chunk.
#
# Get data from previous publication
# https://github.com/cvitolo/GEFF-ERA5/blob/master/data/df_geff_erai_era5.rds
# save it in the "../data" folder
# df <- readRDS(file = "data/df_geff_erai_era5.rds")
# # insert 1-year fire season for tropics (lat = [-30, +30])
# df$season[df$lat < 30 & df$lat > -30] <- "Dry"
# # Remove wet season (we are only interested in the dry season)
# df <- df[df$season == "Dry", ]
# # How many stations remain?
# stations <- unique(df[, c("id", "lat", "long")])
# 
# df$region <- sapply(strsplit(df$tzid, "/"), `[`, 1)
# df$date <- as.Date(paste0(df$yr, "-", df$mon, "-", df$day))
# df <- df[, c("id", "lat", "long", "region", "date", "OBS", "ERA5")]
# 
# df_used <- data.frame(matrix(NA, nrow = 0, ncol = 37))
# names(df_used) <- c("id", "lat", "long", "region", "date", "OBS", "ERA5",
#                     paste0(c("CLIM_d", "HRES_d", "CRPS_d"), rep(1:10, each =3)))
# 
# for (i in seq_along(dates2017)){
# 
#   print(dates2017[i])
# 
#   # What stations have data on this date?
#   dfx <- df %>% filter(date == dates2017[i]) %>% filter(complete.cases(.))
#   # Define spatial point
#   spdf <- dfx[, c("id", "lat", "long")]
#   # Promote data.frame to spatial
#   coordinates(spdf) = ~long + lat
#   crs(spdf) <- new("CRS",
#                    projargs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
#   # Get data from ENS forecasts
#   nstations <- dim(spdf@data)[1]
# 
#   # Get data from HRES forecasts, e.g. ECMWF_FWI_20170101_1200_hr_fwi.nc
#   for (leadtime in 1:10){
# 
#     lead_day <- dates2017[i] + leadtime - 1
# 
#     # Get ERA5 data
#     message(paste("Handling ERA5 for", lead_day))
#     dayx <- which(format(datesERA5, "%m-%d") %in% format(lead_day, "%m-%d")
#                   & lubridate::year(datesERA5) != "2017")
#     system(paste0("cdo seltimestep,", paste0(dayx, collapse = ","),
#                   " /scratch/rd/nen/perClaudia/era5/fwi_1980_2019.nc ",
#                   "/scratch/rd/nen/perClaudia/era5/fwi_1980_2019_dayx.nc"),
#            ignore.stderr = TRUE)
#     ERA5 <- brick("/scratch/rd/nen/perClaudia/era5/fwi_1980_2019_dayx.nc")
#     # ERA5 was stored with longitudes already rotated to [-180, +180],
#     # HRES and ENS will need to be rotated
#     df_era5 <- raster::extract(x = ERA5, y = spdf)
# 
#     # Get HRES
#     message(paste("Handling HRES for", dates2017[i], "day", leadtime))
#     HRES <- raster::brick(file.path(forecasts_folder,
#                                     paste0("ECMWF_FWI_",
#                                            gsub("-", "",
#                                                 as.character(dates2017[i])),
#                                            "_1200_hr_fwi.nc")))[[leadtime]]
#     df_hres <- raster::extract(x = rotate(HRES), y = spdf)
# 
#     # Get ENS
#     df_ens <- matrix(NA, nrow = nstations, ncol = 51)
#     message(paste("Handling ENS for", dates2017[i], "day", leadtime))
#     for (j in 1:51){
#       # Extract the modelled FWI from ENS
#       ens_member <- sprintf("%02d", j - 1)
#       ENS <- raster::brick(file.path(forecasts_folder,
#                                      paste0("ECMWF_FWI_",
#                                             gsub("-", "",
#                                                  as.character(dates2017[i])),
#                                             "_1200_", ens_member,
#                                             "_fwi.nc")))[[leadtime]]
#       df_ens[, j] <- raster::extract(x = rotate(ENS), y = spdf)
#     }
# 
#     # Remove rows with NAs and update spatial points
#     if (leadtime == 1){
#       rows2remove <- which(rowSums(is.na(df_era5)) > 0)
#       if (length(rows2remove) > 0){
#         df_ens <- df_ens[-rows2remove, ]
#         df_hres <- df_hres[-rows2remove]
#         df_era5 <- df_era5[-rows2remove, ]
#         # Resample ERA5 to get 39 random ensemble members
#         df_era5 <- df_era5[, sample(x = 1:dim(df_era5)[2], size = 39,
#                                     replace = TRUE)]
# 
#         dfx <- dfx[-rows2remove, ]
#         # Re-define spatial point
#         spdf <- dfx[, c("id", "lat", "long")]
#         # Promote data.frame to spatial
#         coordinates(spdf) = ~long + lat
#         crs(spdf) <- new("CRS",
#                          projargs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
#         # Get data from ENS forecasts
#         nstations <- dim(spdf@data)[1]
#       }
#     }
# 
#     # Populate CRPS
#     df_crps <- veriApply(verifun = "EnsCrps", fcst = df_ens, obs = df_era5)
#     df_crpss <- apply(X = df_crps, MARGIN = 1, FUN = mean)
#     df_era5_mean <- apply(X = df_era5, MARGIN = 1, FUN = mean)
# 
#     dfx <- cbind(dfx, df_era5_mean, df_hres, df_crpss)
#   }
# 
#   names(dfx) <- names(df_used)
# 
#   # Append to table
#   df_used <- dplyr::bind_rows(df_used, dfx)
#   saveRDS(df_used, "data/df_geff_era5_hres_crps.rds")
# }

df_used <- readRDS("../data/df_geff_era5_hres_crps.rds")

# Promote data.frame to spatial
coordinates(df_used) = ~long + lat

# Add the id column to the GFED4 regions for join
regionXFort <- fortify(BasisRegions, data = BasisRegions@data)
regionXPoly <- merge(regionXFort, BasisRegions@data,
                     by.x = "id", by.y = "ID")  # join data
# Re-order factors
regionXPoly$Region <- factor(as.character(regionXPoly$Region),
                             levels = BasisRegions$Region)

# GFED4 regions of interest
EURO <- BasisRegions[BasisRegions$Region == "EURO",]
TENA <- BasisRegions[BasisRegions$Region == "TENA",]
SHSA <- BasisRegions[BasisRegions$Region == "SHSA",]

# Countries of interest
pt <- raster::getData(name = "GADM", country = "Portugal", level = 1)[-c(2,13),]
chile <- raster::getData(name = "GADM", country = "Chile", level = 0)
cali <- raster::getData(name = "GADM", country = "USA", level = 1)[5, ]

# Synop stations of interest
crs(df_used) <- crs(BasisRegions)
synops <- spatialEco::point.in.poly(df_used, BasisRegions)
# saveRDS(synops, "../data/synops.rds")
```

```{r Figure1, echo = FALSE, fig.width = 13, fig.height = 6, out.width = "100%", fig.cap = "\\label{fig:Figure1}GFED4 regional classification and the 3 countries selected to showcase the fire forecast performances (California, Chile, and Portugal)."}
ggplot(data = map_data("world"), aes(x = long, y = lat)) +
  geom_polygon(aes(group = group), fill=NA, colour = "grey65") +
  coord_equal() +  theme_bw() + xlab("Longitude") + ylab("Latitude") +
  geom_polygon(data = regionXPoly,
               aes(x = long, y = lat, group = group, fill = factor(Region)),
               alpha = 0.5) +
  scale_fill_manual(name="GFED4 regions",
                    values = c("darkgrey",
                               RColorBrewer::brewer.pal(12,"Paired"),
                               "lightgrey"),
                    labels = BasisRegions$Region) +
  geom_polygon(data = map_data("world", region = "Portugal"),
               aes(x = long, y = lat, group = group, colour = "Portugal"),
               fill = NA) +
  geom_polygon(data = map_data("world", region = "Chile"),
               aes(x = long, y = lat, group = group, colour = "Chile"),
               fill = NA) +
  geom_polygon(data = map_data(map = "state", region = "California"),
               aes(x = long, y = lat, group = group, colour = "California"),
               fill = NA) +
  geom_point(data = data.frame(synops), aes(x = coords.x1, y = coords.x2),
             shape = 1, color = "gray20", size = 0.1) +
  scale_colour_manual(name = "Study areas",
                      values = c("Portugal" = "brown",
                                 "California" = "darkblue",
                                 "Chile" = "darkgreen"))
```

```{r Figure2, echo = FALSE, fig.width = 13, fig.height = 10, out.width = "100%", fig.cap = "\\label{fig:Figure3}Comparison between modelled FWI and observed FWI value. FWI are calculated from using ECMWF reanalysis (ERA5) and forecasts at different lead times. The box plots are used to describe the distribution of values accross the observation points for the 2017 Mean Bias (left panel) and the 2017 mean absolute error (right panel)."}
f2 <- data.frame(synops) %>%
  group_by(Region) %>%
  select(OBS, Region, ERA5, HRES_d1, HRES_d2, HRES_d3, HRES_d4, HRES_d5,
         HRES_d6, HRES_d7, HRES_d8, HRES_d9, HRES_d10) %>%
  reshape2::melt(id.vars = c("OBS", "Region")) %>%
  filter(complete.cases(.)) %>% # Remove NAs
  filter(OBS < 250, value < 250) %>%
  mutate(BIAS = OBS - value,
         MAE = abs(OBS - value)) %>%
  rename(modelled_type = variable, modelled_value = value) %>%
  reshape2::melt(id.vars = c("OBS", "Region", "modelled_type",
                             "modelled_value"))

ggplot(f2, aes(x = modelled_type, y = value)) +
  facet_grid(Region ~ variable) +
  geom_boxplot(color="black", outlier.shape = NA) +
  theme_bw() + ylim(-40, 50) +
  xlab("") + ylab("") +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c("ERA5", "D1", "D2", "D3",
                              "D4", "D5", "D6", "D7",
                              "D8", "D9", "D10"))
```

Despite its importance the analysis performed using the synop network is still pointwise and does not cover all the regions where fires are relevant. Moreover mean biases and MAE are based on the high resolution forecast. These skill metrics do not provide information about the performance of the ensemble forecasting as a whole.

```{r Figure3, echo = FALSE, fig.width = 13, fig.height = 6, out.width = "100%", fig.cap = "\\label{fig:Figure4}PUT here crps skills scores."}
f3 <- data.frame(synops) %>%
  select(Region, CRPS_d1, CRPS_d2, CRPS_d3, CRPS_d4, CRPS_d5, CRPS_d6,
         CRPS_d7, CRPS_d8, CRPS_d9, CRPS_d10) %>%
  reshape2::melt(id.vars = "Region") %>%
  filter(complete.cases(.)) %>%
  rename(CRPS = value) %>%
  group_by(Region, variable) %>%
  summarise(CRPS = mean(CRPS))

f3$variable <- as.numeric(f3$variable)
f3$Region <- factor(f3$Region)

ggplot(f3, aes(x = variable, y = CRPS, colour = Region)) +
  geom_line() +
  geom_point() +
  xlab("Lead time") +
  theme_bw() +
  scale_x_discrete(limits = 1:10) +
  scale_colour_manual(name="Regions",
                      values = c("darkgrey",
                                 RColorBrewer::brewer.pal(12,"Paired"),
                                 "lightgrey"),
                      labels = BasisRegions$Region)
```

\subsection{Skill in detecting fire events}

@@

```{r prepFig4, echo = FALSE, message = FALSE, results = 'hide', cache = TRUE}
# Get ERA5
ERA5 <- brick("/scratch/rd/nen/perClaudia/era5/fwi_1980_2019.nc")
# Get FRP
FRP <- raster::brick("/scratch/rd/nen/perClaudia/CAMS/CAMS_2017-01-01_2017-12-31_frpfire.nc")
myfilelist <- list.files(path = forecasts_folder,
                        pattern = "*hr",
                        full.names = TRUE)

library(doSNOW)
ncores <- parallel::detectCores() - 1
cl <- makeCluster(ncores)
registerDoSNOW(cl)
iterations <- 365
pb <- txtProgressBar(max = iterations, style = 3)
progress <- function(n) setTxtProgressBar(pb, n)
opts <- list(progress = progress)
result <- foreach(i = 1:iterations, .combine = rbind, 
                  .options.snow = opts) %dopar%
{
    # Get FRP for 2017, keep only major fires (values above 0.5)
  frpx <- raster::rotate(FRP[[i]])
  p <- raster::rasterToPoints(frpx, fun = function(x){x > 0.5})
  spdf <- data.frame(p)
  spdf$date <- dates2017[i]
  names(spdf) <- c("long", "lat", "frp", "date")
  sp::coordinates(spdf) <- ~long + lat

  # Get HRES for 2017, where FRP is above 0.5
  hres <- raster::rotate(raster::brick(myfilelist[[i]]))
  x <- raster::extract(x = hres, y = spdf)
  x <- data.frame(x)
  names(x) <- paste0("Day", 1:10)

  # Get ERA5 to calculate the very high warning levels
  very_high <- caliver::daily_clima(r = ERA5, dates = dates2017[i], probs = 0.95)
  thr95 <- raster::extract(x = very_high, y = spdf)

  # Append results to table
  df <- cbind(spdf@coords, spdf@data, x, thr95)
  df <- df[complete.cases(df), ]
  return(df)
}
close(pb)
stopCluster(cl)
# saveRDS(result, "data/PODdataframe.rds")

# FRP locations of interest
sp::coordinates(result) <- ~long + lat
crs(result) <- crs(BasisRegions)
fires <- spatialEco::point.in.poly(result, BasisRegions)
# saveRDS(fires, "data/fires.rds")
```

### 2019-12-23 article/article.tex
Article stage after calculating CRPS and extracting data for POD

\subsection{FWI calculation}
\subsubsection{General concept}

@@

The performance of the the fire forecasting systems to reproduce
observed FWI values is assessed using deterministic and probabilistic
scores. Both the synop database and ERA5 are treated as a proxy for
observations in the evaluation. To asses the quality of the computation
we use traditional deterministic skill scores such as the mean bias (MB)
and the mean absolute error (MAE). For a probabilistic assessment, the
continuous ranked probability score is also employed (CRPS;
\cite{hersbach2000}). These are defined as: \begin{align}
MB&=\sum_{p=1}^{cases}\left[F_{n=HRES}-O\right]\\
MAE&=\frac{1}{cases}\sum_{p=1}^{cases}\sqrt{\left[(F_{n=HRES}-O)^2]\right]}\\
CRPS=\frac{1}{cases}\sum_{t=1}^{cases}\int_{-\inf}^{+\inf}\left[F_{n}-O)^2\right]dn
\end{align} where F is the forecast at time step t of N number of
forecasts and O is the observed value. While the MB, MAE and the ACC are
applied to a single forecasts, the high resolution forecast, the CRPS
takes into account the whole distribution of possible values predicted
by the ensemble. It is the continuous extension of the ranked
probability score, where \(F_n\) is the cumulative distribution function
of the predicted ensemble values. Then, the CRPS compares the cumulative
probability distribution of the FWI forecasted by the ensemble forecast
system to the observation. In this sense the CRPS is sensitive to the
mean forecast biases as well as the spread of the ensemble
\citep{hersbach2000}.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{../images/synops_col} \caption{\label{fig:Figure2}Spatial distribution of weather stations from the synop network  which have at least 30 observations recorded at local noon in 2017.}\label{fig:Figure2}
\end{figure}

\begin{verbatim}
## Warning: Removed 958035 rows containing non-finite values (stat_boxplot).
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{article_files/figure-latex/Figure3-1} \caption{\label{fig:Figure3}Comparison between modelled FWI and observed FWI value. FWI are calculated from using ECMWF reanalysis (ERA5) and forecasts at different lead times. The box plots are used to describe the distribution of values accross the observation points for the 2017 Mean Bias (left panel) and the 2017 mean absolute error (right panel).}\label{fig:Figure3}
\end{figure}

@@

\subsection{ Skill in detecting fire events}

>>>>

\subsection{FWI calculation}\subsubsection{General concept}

@@

The performance of the the fire forecasting systems to reproduce
observed FWI values is assessed using deterministic and probabilistic
scores. Both the synop database and ERA5 are treated as a proxy for
observations in the evaluation. To asses the quality of the computation
we use traditional deterministic skill scores such as the mean bias (MB)
and the mean absolute error (MAE). For a probabilistic assessment, the
continuous ranked probability score is also employed (CRPS;
\cite{hersbach2000}). These are defined as:

\begin{align}
MB&=\sum_{p=1}^{cases}\left[F_{n=HRES}-O\right]\\
MAE&=\frac{1}{cases}\sum_{p=1}^{cases}\sqrt{\left[(F_{n=HRES}-O)^2]\right]}\\
CRPS=\frac{1}{cases}\sum_{t=1}^{cases}\int_{-\inf}^{+\inf}\left[F_{n}-O)^2\right]dn
\end{align}

where F is the forecast at time step t of N number of forecasts and O is
the observed value. While the MB, MAE and the ACC are applied to a
single forecasts, the high resolution forecast, the CRPS takes into
account the whole distribution of possible values predicted by the
ensemble. It is the continuous extension of the ranked probability
score, where \(F_n\) is the cumulative distribution function of the
predicted ensemble values. Then, the CRPS compares the cumulative
probability distribution of the FWI forecasted by the ensemble forecast
system to the observation. In this sense the CRPS is sensitive to the
mean forecast biases as well as the spread of the ensemble
\citep{hersbach2000}.

@@

\begin{verbatim}
## Warning: Removed 104425 rows containing non-finite values (stat_boxplot).
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{article_files/figure-latex/Figure2-1} \caption{\label{fig:Figure3}Comparison between modelled FWI and observed FWI value. FWI are calculated from using ECMWF reanalysis (ERA5) and forecasts at different lead times. The box plots are used to describe the distribution of values accross the observation points for the 2017 Mean Bias (left panel) and the 2017 mean absolute error (right panel).}\label{fig:Figure2}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{article_files/figure-latex/Figure4-1} \caption{\label{fig:Figure4}PUT here crps skills scores.}\label{fig:Figure4}
\end{figure}

\subsection{Skill in detecting fire events}
